MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-05-23 MAESTROeX_I/algorithm.tex
some tweaks

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry. The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the radial array cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the radial array cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of a Cartesian grid variable onto a one-dimensional radial array.
In planar geometries, this is a straightforward arithmetic average of cells at
a particular height since the radial cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any Cartesian cell-center can map to.  
Specifically, for every Cartesian cell, there exists an integer $m$ such that the radius to the center of the star is given by $\hat{r}_m=\Delta x\sqrt{0.75+2m}$.
We average by binning all the cells associated with each radii and taking the average.
Then we interpolate this data onto a one-dimensional radial array.  
Previously, MAESTRO only allowed for radial arrays with a constant $\Delta r$ (typically equal to $\Delta x/5$).
Here we present a new option to retain an irregularly-spaced radial array to eliminate mapping errors back onto the Cartesian grid state.
Consider a spherical star in hydrostatic equlibrium at rest.  In the absense of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced array, the effects due to interpolation error are completely eliminated.
\MarginPar{Doreen you had a great figure of this mapping in one of your talks}

@@

At the beginning of each time step we have the cell-centered state,
$(\Ub,\rho X_k,\rho h,\rho_0,p_0)^n$, and nodal state, $\pi^{n-\myhalf}$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state, e.g.,
\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed planar volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution; see \cite{XRB_I}.  We are still exploring this option for spherical stars.}
\begin{equation}
T = T(\rho,p_0,X_k),
\end{equation}
and ($\gammaonebar,\beta_0)$ are computed from $(\rho,p_0,X_k)$ (see Appendix A of Paper I and Appendix C of Paper III for details).

The overall flow of the algorithm is a second-order Strang splitting approach for the coupling of the reactions and advection of the thermodynamic variables.  
We use a predictor-corrector approach within the Strang splitting scheme to achieve second-order accuracy in time.
We integrate the velocity using a standard second-order projection method to enforce the divergence constraint.
To summarize:
\begin{itemize}
\item In {\bf Step 1} we react the thermodynamic variables over $\Delta t/2$.
\item In {\bf Steps 2-4} we advect the thermodynamic variables over $\Delta t$.  Specifically, we compute an estimate for the expansion term, $S$, project the face-centered, time-centered velocities so they satisfy the divergence constraint, and then advect the thermodynamic variables.
\item In {\bf Step 5} we react the thermodynamic variables over $\Delta t/2$.\footnote{After this step we could skip to the velocity advance in {\bf Steps 10-11}, however the overall scheme would be only first-order in time, so {\bf Steps 6-9} can be thought of as a trapezoidal corrector step.}
\item In {\bf Steps 6-8} we redo the advection in {\bf Steps 2-4} but are able to use the trapezoidal rule to time-center certain quantities such as $S$, $\rho_0$, etc.
\item In {\bf Step 9} we redo the reactions from {\bf Step 5} using the improved results for the corrector advection step.
\item In {\bf Steps 10-11} we and update the velocity, compute the expansion term, $S$, and project the cell-centered velocities.
\end{itemize}

@@

In stratified low Mach number models, due to the extreme variation in density, the velocity can become very large in low density regions at the edge of the star.
Throughout Papers II-V, we have employed two techniques to help control these dynamics without significantly affecting the dynamics in the convective region.
First, we use a cutoff density technique, where we hold the density constant outside a specified radius (typically near where the density is $\sim$4 orders of magnitude smaller than the largest densities in the simulation.
Second, we employ a sponge technique where we artificially damp the velocities near and beyond the cutoff region.
For more details, refer to Paper V and the previous references cited within.

>>>>

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry. The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids. \MarginPar{did this fig come from one of the earlier papers? if so, we will need to acknowledge that paper and make sure ApJ knows.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the radial array cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the radial array cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of a Cartesian grid variable onto a one-dimensional radial array.
In planar geometries, this is a straightforward arithmetic average of cells at
a particular height since the radial cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any Cartesian cell-center can map to.  
Specifically, for every Cartesian cell, there exists an integer $m$ such that the radius to the center of the star is given by $\hat{r}_m=\Delta x\sqrt{0.75+2m}$.
We average by binning all the cells associated with each radii and taking the average.
Then we interpolate this data onto a one-dimensional radial array.  
Previously, MAESTRO only allowed for radial arrays with a constant $\Delta r$ (typically equal to $\Delta x/5$).
Here we present a new option to retain an irregularly-spaced radial array to eliminate mapping errors back onto the Cartesian grid state.
Consider a spherical star in hydrostatic equlibrium at rest.  In the absense of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced array, the effects due to interpolation error are completely eliminated.
\MarginPar{Doreen you had a great figure of this mapping in one of your talks}

@@

At the beginning of each time step we have the cell-centered state,
$(\Ub,\rho X_k,\rho h,\rho_0,p_0)^n$, and nodal state, $\pi^{n-\myhalf}$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed planar volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution; see \cite{XRB_I}.  We are still exploring this option for spherical stars.}, e.g.,
\begin{equation}
T = T(\rho,p_0,X_k),
\end{equation}
and ($\gammaonebar,\beta_0)$ are computed from $(\rho,p_0,X_k)$ (see Appendix A of Paper I and Appendix C of Paper III for details).

The overall flow of the algorithm is a second-order Strang splitting approach for the coupling of the reactions and advection of the thermodynamic variables.  
We use a predictor-corrector approach within the Strang splitting scheme to achieve second-order accuracy in time.
We integrate the velocity using a standard second-order projection method to enforce the divergence constraint.
To summarize:
\begin{itemize}
\item In {\bf Step 1} we react the thermodynamic variables over $\Delta t/2$.
\item In {\bf Steps 2--4} we advect the thermodynamic variables over $\Delta t$.  Specifically, we compute an estimate for the expansion term, $S$, project the face-centered, time-centered velocities so they satisfy the divergence constraint, and then advect the thermodynamic variables.
\item In {\bf Step 5} we react the thermodynamic variables over $\Delta t/2$\footnote{After this step we could skip to the velocity advance in {\bf Steps 10--11}, however the overall scheme would be only first-order in time, so {\bf Steps 6-9} can be thought of as a trapezoidal corrector step.}.
\item In {\bf Steps 6--8} we redo the advection in {\bf Steps 2--4} but are able to use the trapezoidal rule to time-center certain quantities such as $S$, $\rho_0$, etc.
\item In {\bf Step 9} we redo the reactions from {\bf Step 5} using the improved results for the corrector advection step.
\item In {\bf Steps 10--11} we and update the velocity, compute the expansion term, $S$, and project the cell-centered velocities.
\end{itemize}

@@

In stratified low Mach number models, due to the extreme variation in density, the velocity can become very large in low density regions at the edge of the star.
Throughout Papers II-V, we have employed two techniques to help control these dynamics without significantly affecting the dynamics in the convective region.
First, we use a cutoff density technique, where we hold the density constant outside a specified radius (typically near where the density is $\sim$4 orders of magnitude smaller than the largest densities in the simulation).
Second, we employ a sponge technique where we artificially damp the velocities near and beyond the cutoff region.
For more details, refer to Paper V and the previous references cited within.
